The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-24 paper.tex
Restricao de "delayed awaits"

\begin{comment}
\vspace{5pt}
\CEU provides no support for standard functions for a number of reasons:
\begin{itemize}
\item The interaction with other \CEU control primitives is not obvious (e.g., 
executing an $await$ or a $par/or$ inside a function).
\item They would still be restricted in some ways given the embedded context 
(e.g.  no recursion or closures).
\item Programs can always recur to $C$ functions for low-level operations.
%\item A dedicated primitive would behave just as described, being a matter of 
%syntactic sugar.
\end{itemize}

Regardless of the limitations, this form of subroutines is widely adopted in 
\CEU programs, given that they were designed to work with the other control 
mechanisms.
Keep in mind that the typical reactive organization of programs (awaiting an 
external stimulus, reacting to it, and going back to awaiting) does not demand 
unrestricted subroutines.
\end{comment}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
%\newline
{\small
%TODO.
}
\label{lst.shared}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
{\small
%TODO.
}
\label{lst.det}
}
\end{figure}

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}
%
At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ will prevent that awaiting expressions 
awake in the same reaction they are reached (ensuring that the program can 
progress).

>>>>

Another distinction regarding event handling is that Esterel supports 
same-cycle bi-directional communication~\cite{esterel.compiling}, i.e., two 
threads can react to one another during the same cycle due to mutual signal 
dependency.
%
\CEU imposes a restriction that an \code{await} is only valid for the next 
reaction, i.e., if an \code{await} and \code{emit} for the same event occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements})~\cite{esterel.schizo}.
%
The example of Figure~\ref{lst.delay} relies on this restriction to avoid 
infinite execution:
Although both sides of the \code{par/or} have an \code{await} statement, the 
\code{loop} would restart instantaneously if the \code{await} could awake in 
the same reaction that reaches it.
%
In an atypical scenario requiring immediate awake, delayed awaits can be 
circumvented by judiciously copying the code to execute on awake.

\begin{figure}
\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;
        await f;
    end
end
\end{lstlisting}
\caption{ The \code{await} occurs simultaneously with the \code{emit} and does 
not awake, preventing the \code{par/or} to terminate and execute the 
\code{loop} indefinitely.
\label{lst.delay}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
[
    call f1();
$\|$
    call f2();
];
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
par/and do
    _f1();
with
    _f2();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2}.
\label{lst.det}
}
\end{figure}

@@

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}
%
At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ prevents that awaiting expressions 
awake in the same reaction they are reached (the \emph{delayed awaits} as 
explained in Section~\ref{sec.ceu.ints}).
